Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-27 paper.tex
Various small edits
Intro parts

\author[mcgill]{Reza Adalat}
\author[mcgill]{Natacha Beck}
\author[mcgill]{Samir Das}
\author[isi]{Rafael Ferreira da Silva}
\author[mcgill]{Najmeh Khalili-Mahani}
\author[criugm]{Pierre-Olivier Quirion}
\author[mcgill]{Pierre Rioux}

@@

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular clusters, to handle high-throughput
task executions. For instance, FSL can launch tasks on SGE through its
fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These
domain-specific engines represent a tremendous opportunity for science
gateways to leverage existing tools and applications. They nicely
complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

For the sake of the analysis conducted in this paper, we define a
\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science and
domain-specific engines. Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources within a community and to
provide increased performance and capacity through facilitated access
to storage and computing power. They are often accessible through a
web interface that helps users manage access rights, data transfers,
task execution and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways may be tailored to a particular
engine that manages critical functions such as task execution, data
transfers and authentication. Other gateways may be more general and
host applications executed by different types of engines. Finally,
some science gateways may not use any workflow engine and focus on the
execution of simple tasks -- this class of systems is not covered by
our analysis.

Extensibility is an important property of the integration, as new
workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted.

Scalability is also a crucial concern, since science gateways are
multi-user systems that manage substantial amounts of data and execute
computationally-intensive analyses. To this end, science gateways
may include different instances of the same engine type and implement
some load-balancing policy among them. They may also use advanced
task scheduling policies on the infrastructure, e.g. to ensure fairness
among users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble science gateways. The main
frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
were built using such
frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
instances were also built independently from any framework, using
regular web development tools and libraries to access distributed
systems. CBRAIN and VIP are two such science gateways, from which we
gathered experience about the architectures described in this paper.

CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
open-source\footnote{\url{https://github.com/aces/cbrain}}, web-based collaborative platform addressing the
challenges of data-heavy, computation-intensive neuroimaging
research. It offers transparent web-based access to remote data
sources, distributed computing sites, and an array of processing and
visualization tools, within a secure environment. CBRAIN is deployed
at a wide range of computing sites, including a network of national HPC
centers in Canada, one in Korea, one in Germany, and on numerous local
servers. As February 2016, CBRAIN has served over 420 users across 50
cities in 22 countries and has a national computing allocation of 660 CPU years
per annum.

@@

In this paper, infrastructure consists of the computing and storage
resources involved in the workflow execution, as well as the software
services used to access these resources. Infrastructures can be composed of
computing or file servers, databases, clusters, grids or clouds. Some
workflow engines and science gateways may assume specific
characteristics about the infrastructure, such as the presence of a
shared file system between the computing nodes, the availability of a
global task meta-scheduler, the presence of a file catalog, etc. In
the analysis presented below, such specific functions are not
discussed. Instead, we consider the infrastructure as an abstract
system that can execute tasks and store data regardless of the
enabling mechanisms.

@@

New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99DD99}4
                                     & \cellcolor[HTML]{99DD99}4
                                     & \cellcolor[HTML]{99FF99}3
                                     & \cellcolor[HTML]{99DD99}4
                                     & \cellcolor[HTML]{99DD99}4\\
  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D199}\textbf{10}
                                     & \cellcolor[HTML]{99D199}\textbf{10}
                                     & \cellcolor[HTML]{99FF99}\textbf{8}
                                     & \cellcolor[HTML]{999999}\textbf{12.5}
                                     & \cellcolor[HTML]{99E899}\textbf{9}\\

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is also a consequence of global
complexity and it is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain). 
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to modify the science gateway as well as interactions
\texttt{b} and \texttt{c$_2$} (\texttt{E$_1$}=3). Updating a workflow
engine version requires modifications in the science gateway
(\texttt{E$_2$}=1).  Inserting a new workflow is done through
interactions \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$}=2). Adding a new infrastructure generates updates in
interactions \texttt{a$_1$}, \texttt{b}, \texttt{c$_1$} and
\texttt{c$_2$} (\texttt{E$_4$}=4).

@@

\paragraph{Extensibility} Adding a new type of workflow engine
requires to implement the corresponding workflow service, to modify
interaction \texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction or
component (\texttt{E$_2$}=0). Updating the engine version in a
workflow service does not count since the only goal of this component
is to wrap the engine.  New workflows are added in the science gateway
or in the workflow engine through interactions \texttt{a$_1$} and
\texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new type of infrastructure
requires updates in interactions \texttt{a$_1$}, \texttt{b},
\texttt{$c_1$} and \texttt{$c_2$} (\texttt{E$_4$}=4).

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
requires updates only on the infrastructure (\texttt{E$_2$}=1).  New
workflows are integrated by creating a new task in the science gateway
through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
infrastructure requires to update interactions \texttt{a$_1$}, \texttt{b} and
\texttt{c} in the science gateway (\texttt{E$_4$}=3).

@@

\paragraph{Extensibility} Adding a new engine type requires to wrap
the engine in the agent and to update interactions \texttt{b} and
\texttt{c$_2$} (\texttt{E$_1$=3}). Updating the version of an engine
is transparent (\texttt{E$_2$=0}) since it only requires updating the
agent that is a component dedicated to the engine. Integrating a new workflow is
done through interactions \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$=2}). Integrating a new infrastructure requires updates
in interactions \texttt{a$_1$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{E$_4$=4}).

@@

\paragraph{Extensibility} Adding a new type of \emph{parent} engine
requires to implement the corresponding service, to implement
interactions \texttt{b} and \texttt{c$_2$} in the parent engine, and
to implement interaction \texttt{d} in the science gateway and in the
parent service (\texttt{E$_1$=5}). Adding a new type of \emph{child}
engine only requires to implement the corresponding service, to
develop interactions \texttt{b} and \texttt{c$_2$} in the child
engine, and to implement interaction \texttt{d} in the parent service
(\texttt{E$_1$=4}). We use \texttt{E$_1$=4.5} in
Table~\ref{table:evaluation} to reflect both conditions. Adding a new
version in the parent or child engine only requires modifying this
engine (\texttt{E$_2$=0}). Adding a new
workflow is done through interaction \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$=2}). Adding a new infrastructure requires to
re-implement interactions \texttt{a$_1$}, \texttt{b} and \texttt{c$_2$} twice, and
interaction \texttt{c$_1$} once (\texttt{E$_4$=6}).

@@

\paragraph{Extensibility} Since adding a new type of workflow engine
aims at supporting more workflows, we consider that it only requires
to re-implement interaction \texttt{g} in this architecture  (\texttt{E$_1$=1}). Note,
however, that implementing interaction \texttt{g} can require very
substantial work depending on the complexity of the language used by
the new engine. Based on the same logic, adding a
new engine version only requires modifying interaction \texttt{g}
(\texttt{E$_2$=1}).  Adding a new workflow is done through
interactions \texttt{a$_1$}, \texttt{a$_2$} and \texttt{g} (\texttt{E$_3$=3}). As in the
service architecture, interfacing with a new infrastructure requires
modifications in the workflow service and in interactions \texttt{a$_1$}, \texttt{b}
and \texttt{c$_2$} (\texttt{E$_4$=4}).

>>>>

\author[mcgill]{Reza Adalat}
\author[mcgill]{Natacha Beck}
\author[isi]{Rafael Ferreira da Silva}
\author[mcgill]{Najmeh Khalili-Mahani}
\author[criugm]{Pierre-Olivier Quirion}
\author[mcgill]{Pierre Rioux}

@@

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing systems, in particular cluster schedulers, to handle high-throughput
task executions. For instance, FSL can launch tasks on SGE through its
fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These
domain-specific engines represent a tremendous opportunity for science
gateways to leverage existing tools and applications. They nicely
complement e-Science engines that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

For the sake of the analysis conducted in this paper, we define a
\emph{workflow engine} (also abbreviated \emph{engine}) as a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\note{Marc-e}{To remain consistent with above descriptions} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Science gateways are used to share resources in a community and to
provide increased performance and capacity through facilitated access to storage and
computing power. They are often accessible through a web interface
that helps users manage access rights, data transfers, task execution
and authentication on multiple computing and storage
locations. Workflow engines are part of this ecosystem as core
components to implement and execute applications.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways may be tailored to a particular
engine that manages critical functions such as task execution, data
transfers and authentication. Other gateways may be more general and
host applications executed by different types of
engines. Extensibility is an important property of the integration, as
new workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted.

Scalability is also an important concern, since science gateways are
multi-user systems that manage substantial amounts of data and execute
computationally-intensive analyses. To this end, science gateways
may include different instances of the same engine type and implement
some load-balancing policy among them. They may also use advanced
task scheduling policies on the infrastructure, e.g. to ensure fairness
among users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble science gateways. The main
frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
were built using such
frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
instances were also built independently from such specialized framework, using
regular web development tools and libraries to access distributed
systems. CBRAIN and VIP are two such science gateways, from which we
gathered experience about the architectures described in this paper.

CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
open-source\footnote{\url{https://github.com/aces/cbrain}}, web-based collaborative platform addressing the
challenges of data-heavy, computation-intensive neuroimaging
research. It offers transparent web-based access to remote data
sources, distributed computing sites, and an array of processing and
visualization tools, within a secure environment. CBRAIN is deployed
at a wide range of computing sites, including a network of national HPC
centers in Canada, a cluster in Germany, and on numerous local research
servers. CBRAIN is a task-centric platform where various workflow engines
can be deployed on the infrastructure as any other task would. As February 2016,
CBRAIN has served over 420 users across 50 cities in 22 countries and has a
national computing allocation of 660 CPU years per annum.

@@

In this paper, infrastructure consists of the computing and storage
resources involved in the workflow execution, as well as the software
services used to access these resources. Infrastructures can be composed of
computing or file servers, databases, clusters, grids or clouds. Some
workflow engines and science gateways may assume specific
characteristics about the infrastructure, such as the presence of a
shared file system between the computing nodes, the availability of a
global task meta-scheduler, the presence of a file catalog, etc. In
the analysis presented below, such specific requirements are not
discussed. Instead, we consider the infrastructure as an abstract
system that can execute tasks and store data regardless of the
enabling mechanisms.

@@

New infrastructure -- \texttt{E$_4$} & \cellcolor[HTML]{99DD99}3
  \textbf{Total} (difficulty to extend) & \cellcolor[HTML]{99D199}\textbf{9}
                                     & \cellcolor[HTML]{99D199}\textbf{9}
                                     & \cellcolor[HTML]{99E899}\textbf{8}
                                     & \cellcolor[HTML]{999999}\textbf{11.5}
                                     & \cellcolor[HTML]{99E899}\textbf{8}\\

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is also a consequence of global
complexity and it is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to modify the science gateway as well as interactions
\texttt{b} and \texttt{c$_2$} (\texttt{E$_1$}=3). Updating a workflow
engine version requires modifications in the science gateway
(\texttt{E$_2$}=1).  Inserting a new workflow is done through
interactions \texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new infrastructure generates updates in
interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{E$_4$}=3).

@@

\paragraph{Extensibility} Adding a new type of workflow engine
requires to implement the corresponding workflow service, to modify
interaction \texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{E$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction or
component (\texttt{E$_2$}=0). We consider that updating the engine
version in a workflow service does not require code updates in the
service since the only goal of this component is to wrap the engine.
New workflows are added in the science gateway or in the workflow
engine through interactions \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$}=2). Adding a new type of infrastructure requires
updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
(\texttt{E$_4$}=3).

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
requires updates only on the infrastructure (\texttt{E$_2$}=1).  New
workflows are integrated by creating a new task in the science gateway
through interactions \texttt{a$_1$} and \texttt{a$_2$}  (\texttt{E$_3$}=2). Adding a new
infrastructure requires to update interactions \texttt{b} and
\texttt{c} in the science gateway (\texttt{E$_4$}=2).

@@

\paragraph{Extensibility} Adding a new engine type requires to wrap
the engine in the agent and to update interactions \texttt{b} and
\texttt{c$_2$} (\texttt{E$_1$=3}). Updating the version of an engine
is transparent (\texttt{E$_2$=0}) since it only requires updating the
agent that is a component dedicated to the engine. Integrating a new workflow is
done through interactions \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$=2}). Integrating a new infrastructure requires updates
in interactions \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{E$_4$=3}).

@@

\paragraph{Extensibility} Adding a new type of \emph{parent} engine
requires to implement the corresponding service, to implement
interactions \texttt{b} and \texttt{c$_2$} in the parent engine, and
to implement interaction \texttt{d} in the science gateway and in the
parent service (\texttt{E$_1$=5}). Adding a new type of \emph{child}
engine only requires to implement the corresponding service, to
develop interactions \texttt{b} and \texttt{c$_2$} in the child
engine, and to implement interaction \texttt{d} in the parent service
(\texttt{E$_1$=4}). We use \texttt{E$_1$=4.5} in
Table~\ref{table:evaluation} to reflect both conditions. Adding a new
version in the parent or child engine only requires modifying this
engine (\texttt{E$_2$=0}). Adding a new
workflow is done through interaction \texttt{a$_1$} and \texttt{a$_2$}
(\texttt{E$_3$=2}). Adding a new infrastructure requires to
re-implement interactions \texttt{b} and \texttt{c$_2$} twice, and
interaction \texttt{c$_1$} once (\texttt{E$_4$=5}).

@@

\paragraph{Extensibility} Since adding a new type of workflow engine
aims at supporting more workflows, we consider that it only requires
to re-implement interaction \texttt{g} in this architecture  (\texttt{E$_1$=1}). Note,
however, that implementing interaction \texttt{g} can require very
substantial work depending on the complexity of the language used by
the new engine. Based on the same logic, adding a
new engine version only requires modifying interaction \texttt{g}
(\texttt{E$_2$=1}).  Adding a new workflow is done through
interactions \texttt{a$_1$}, \texttt{a$_2$} and \texttt{g} (\texttt{E$_3$=3}). As in the
service architecture, interfacing with a new infrastructure requires
modifications in the workflow service and in interactions \texttt{b}
and \texttt{c$_2$} (\texttt{E$_4$=3}).
